---
title: "00_data_preparation"
author: "Lennert Schepers"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("DT")) install.packages('DT')
library(DT)
```

##Data preparation

This shows the code that was run to prepare the data for the Machine Learning workshop at the OpenSeaLab Hackathon 2019.
Note that for some steps you need to download the data manually.

* download EMODnet Biology data
* download Substrate

However, you can do this also automatically by using webservices. We will update this document and include these automatic steps.


### EMODnet Biology data

We are using the 'Dutch national shellfish monitoring in the coastal zone':
http://www.emodnet-biology.eu/data-catalog?module=dataset&dasid=6204


See selection here, click 'download'
<http://www.emodnet-biology.eu/toolbox/en/download/selection/15d69183543357>

```{r load_substrate, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
library(sf)
library(mapview)
library(raster)

# data downloaded in /data/20190830_162455_15d6931b78ad5a.zip
unzip("data/20190830_162455_15d6931b78ad5a.zip",
       exdir = "data")

# read csv
Dutch_shell_full <- read.csv("data/DATA_63zArr.csv")

# convert to sf object
Dutch_shell_sf <- st_as_sf(Dutch_shell_full,
                           coords = c('decimallongitude','decimallatitude'),
                           crs = 4326,
                           remove = FALSE)

# convert date-time column to POSIXct class
Dutch_shell_sf$datecollected <- as.POSIXct(Dutch_shell_sf$datecollected,
                                           format="%Y-%m-%dT%H:%M")

# viz first 50 points on the map
mapview(Dutch_shell_sf[1:50,],
        zcol = "yearcollected")
```

We will select different data other parameters for our measurements, therefore we need a bounding box of the measurements:
```{r bbox, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
# determine min max coordinate values
# xmin <- st_bbox(Dutch_shell_sf)$xmin
# ymin <- st_bbox(Dutch_shell_sf)$ymin
# xmax <- st_bbox(Dutch_shell_sf)$xmax
# ymax <- st_bbox(Dutch_shell_sf)$ymax
bbox <- st_bbox(Dutch_shell_sf)

```


### Additional variables
#### Fishing vessel density

We want to retrieve the ship vessel density data (especially Fishing vessels) from EMODnet Human Activities.
Have a look to the EMODnet Human Activities WCS service (raster layers)

<https://ows.emodnet-humanactivities.eu/wcs?service=wcs&version=1.0.0&request=GetCapabilities>

This returns a xml with all coverages, for example:

```
<wcs:ContentMetadata>
  <wcs:CoverageOfferingBrief>
    <wcs:description>Generated from GeoTIFF</wcs:description>
    <wcs:name>emodnet:2017_01_st_00</wcs:name>
    <wcs:label>2017_01_st_00</wcs:label>
  ...
</Layer>
```
In the example above we see the desciption of the coverage of

* year 2017
* month 01 (January)
* shiptype 00 ('Other')

See the table below for all options

Year_month | ship type                                        
------------------------| -----------------------------------------------
2017_01 | st_00: Other 
2017_02 | st_01: Fishing 
2017_03 | st_02: Service 
2017_04 | st_03: Dredging or underwater ops 
2017_05 | st_04: Sailing |  
2017_06 | st_05: Pleasure Craft  
2017_07 | st_06: High speed craft  
2017_08 | st_07: Tug and towing  
2017_09 | st_08: Passenger 
2017_10 | st_09: Cargo  
2017_11 | st_10: Tanker
2017_12 | st_11: Military and Law Enforcement 
...        | st_12: Unknown 
Year average: 2017_ [st_..]_avg  | st_All: All ship types 


some examples of layers:

Description | wcs coverage name                                       
------------------------| -------
average of 2017 Sailing density: | `emodnet:2017_st_04_avg`
average of October Dredging: | `emodnet:2017_10_st_03`
average of 2017 all types: | `emodnet:2017_st_00_avg`


Let's extract the 2017 average of Fishing:

```{r vesseldensity, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
coveragename <- 'emodnet:2017_st_01_avg'

url <- 'https://ows.emodnet-humanactivities.eu/wcs?service=wcs&version=1.0.0&request=getcoverage'
full_url <- paste(url,
                  '&coverage=', coveragename,
                  '&crs=EPSG:4326&BBOX=', paste(bbox, collapse = ","),
                  '&format=image/tiff',
                  '&interpolation=nearest&resx=0.00833333&resy=0.00833333',
                  sep = '')

fishing <- raster(full_url)
mapview(fishing)

```

#### EMODnet Bathymetry
Have a look to the EMODnet Bathymetry WCS service (raster layers)

<https://ows.emodnet-humanactivities.eu/wcs?service=wcs&version=1.0.0&request=GetCapabilities>

This returns a xml with all coverages, for example:

```
<wcs:ContentMetadata>
  <wcs:CoverageOfferingBrief>
    <wcs:description>Generated from GeoTIFF</wcs:description>
    <wcs:name>emodnet:2017_01_st_00</wcs:name>
    <wcs:label>2017_01_st_00</wcs:label>
  ...
</Layer>
```

Let's extract the mean bathymetry:

```{r bathymetry, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
coveragename <- 'emodnet:mean'

url <- 'https://ows.emodnet-bathymetry.eu/wcs?service=wcs&version=1.0.0&request=getcoverage'
full_url <- paste(url,
                  '&coverage=', coveragename,
                  '&crs=EPSG:4326&BBOX=', paste(bbox, collapse = ","),
                  '&format=image/tiff',
                  '&interpolation=nearest&resx=0.00833333&resy=0.00833333',
                  sep = '')

bathy <- raster(full_url)
mapview(bathy)

```

Extract rasterdata at locations:

```{r rasterextract, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
Dutch_shell_sf$fishing <- extract(fishing, Dutch_shell_sf)
Dutch_shell_sf$bathy <- extract(bathy, Dutch_shell_sf)
```


#### Substrate


Download substrate data from:
<https://www.emodnet-seabedhabitats.eu/access-data/download-data/>
We downloaded the 'EUSeaMap → Classified habitat descriptors → Substrate type'

```{r substrate, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
library(sf)
library(raster)
library(ggplot2)
library(mapview)
library(fasterize) #https://github.com/ecohealthalliance/fasterize
# unzip file:
# unzip('data/eusm2019_substrate.zip',
#       exdir = 'data')

 # try with substrate
 substrate <- st_read(dsn = 'data/EUSM2019_Substrate/EUSM2019_substrate.gdb',
                      layer = 'EUSM2019_Arctic_Atlantic_substrate')
 st_crs(substrate)

# subset data by North Sea:
  # bounding box North Sea
  # xmin, ymin, xmax, ymax
  # -4.45, 51.00, 12.01, 61.02
  NS <- st_polygon(list(rbind(c(-4.45,51.00),
                                 c(12.01,51.00),
                                 c(12.01,61.02),
                                 c(-4.45,61.02),
                                 c(-4.45,51.00))))
  NS <- st_sfc(NS) %>% st_set_crs(NA) %>% st_set_crs(4326)
  
  # transform coordinates to match habitats CRS
  NS <- NS %>% st_transform(3857)

  # cut (intersect) substrate by North Sea bbox
  NS_substrate <- st_intersection(substrate, NS)

 # the Seabed substrate map originates from a very detailed vector layer
# so convert polygons to raster on at aggregated resolution
# raster template
r_template <- raster(ncols = 791, nrows = 481, 
                     crs = projection(NS_substrate), 
                     ext = extent(NS_substrate),
                     vals = NULL)

# apparently not all polygon type (error with fasterize)
NS_substrate <- st_cast(NS_substrate, "POLYGON")

# Fasterize for amazingly fast sf to raster! see https://github.com/ecohealthalliance/fasterize
r_NS_substrate_f <- fasterize(NS_substrate, r_template, field = "Substrate")

# # convert to dataframe for ggplot
r_NS_Substrate_spdf <- as(r_NS_substrate_f, 'SpatialPixelsDataFrame')
r_NS_Substrate_df <- as.data.frame(r_NS_Substrate_spdf)
colnames(r_NS_Substrate_df) <- c('Substrate', 'lon', 'lat')
# 
# plot(r_NS_Substrate, axes = TRUE)
# 
# # Plot data in ggplot
ggplot() +
  geom_tile(data=r_NS_Substrate_df, aes(x=lon, y=lat, fill=factor(Substrate))) +
  scale_fill_viridis_d(breaks = as.character(1:length(levels(NS_substrate$Substrate))),
                       labels = levels(NS_substrate$Substrate),
                       name = 'Substrate')

```

Extract substrate data from rasters:

```{r extractsubstrate, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
Dutch_shell_sf$substrate <- extract(r_NS_substrate_f , Dutch_shell_sf)

# add levels to substrate:
Dutch_shell_sf$substrate <- factor(Dutch_shell_sf$substrate,
                                   levels = 1:length(levels(NS_substrate$Substrate)),
                                   labels = levels(NS_substrate$Substrate))
```

#### Bio-Oracle layers

We use other descriptors from the sdmpredictors package:

<http://www.bio-oracle.org/>

sdmpredictors contains several datasets, each containing several layers:
```{r biooracle, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
library(sdmpredictors)
# list datasets of sdmpredictors package
datatable(list_datasets())

# list layers of Bio-ORACLE dataset
datatable(list_layers(datasets="Bio-ORACLE"))
```

load all interesting layers:

```{r , message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
# load some Bio-Oracle layers:
bo_layers <- load_layers(layercodes = c("BO_phosphate",
                           "BO_nitrate",
                           "BO_salinity",
                           "BO_sstmax",
                           "BO_sstmean",
                           "BO_dissox",
                           "BO2_curvelmax_bdmax",  # Current velocity (maximum at max depth)
                           "BO2_lightbotmax_bdmax"), # Light at bottom (maximum at max depth)
                         rasterstack = TRUE )

Dutch_shell_sf <- cbind(Dutch_shell_sf,
                        extract(bo_layers, Dutch_shell_sf))

```


#### Distance to the coast

Get coastline from rnaturalearth package:
```{r rnaturalearth, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
library(rnaturalearth)
BENLGER <- ne_countries(scale = 'large',
                     country = c('Belgium', 'Netherlands', 'Germany'),
                     type = 'countries')
BENLGER <- st_as_sf(BENLGER) %>% st_combine()

mapview(BENLGER)

```
Calculate distance to the coastline:

```{r, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
# Calculate distance to the coastline:
coastdist <- st_distance(BENLGER, Dutch_shell_sf)

# Combine to dataframe:
Dutch_shell_sf$coastdist <- as.vector(coastdist)
  
```

Check if it looks ok (for first 50 records):

```{r, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
mapview(Dutch_shell_sf[1:50,],
        zcol = "coastdist")
```



Select only a selection of the parameters:

```{r, message=FALSE, warning=FALSE, error=FALSE, fig.align='center'}
library(dplyr)

Dutch_shell_sel <- Dutch_shell_sf %>%
select(datecollected, decimallongitude, decimallatitude,
       aphiaid, scientificnameaccepted,
       kingdom, phylum, class,
       order, family, genus,
       subgenus, specificepithet, infraspecificepithet,
       Bedabundance = BedAbund....m.2.,
       BedWetWtBiom = BedWetWtBiom..g.m.2.,
       fishing, bathy, substrate,
       BO_phosphate, BO_nitrate, BO_salinity,
       BO_sstmax, BO_sstmean,
       BO_dissox, BO2_curvelmax_bdmax,
       BO2_lightbotmax_bdmax, geometry, coastdist)
write.csv(Dutch_shell_sel,
          file = "data/Dutch_shell_variables.csv",
          row.names = FALSE,
          quote = grep('geometry', colnames(Dutch_shell_sel))
)
write.table(Dutch_shell_sel,
          file = "data/Dutch_shell_variables_semicolon.csv",
          row.names = FALSE,
          sep = ";")

```

So in the end we have a dataset with following variables:

* datecollected, decimallongitude, decimallatitude: *time + coordinates*
* aphiaid, scientificnameaccepted, kingdom, phylum, class, order, family, genus, subgenus, specificepithet, infraspecificepithet: *taxonomic variables*
* Bedabundance: *individuals per square meter (#/m²)*
* BedWetWtBiom : *Wet weight biomass of biological entity (g/m²)*
* Fishing: *fishing intensity* (average number of hours spent by Fishing ships in a square kilometre over a month)
* Bathy: *bathymetry*
* substrate: *seabed substreate*
* BO_phosphate, BO_nitrate, BO_salinity: *environmental layers: phosphate, nitrate, salinity*
* BO_sstmax, BO_sstmean: : *max and mean sea surface temperature*
* BO_dissox: *dissolved oxygen*
* BO2_curvelmax_bdmax: *Current velocity (maximum at max depth)*
* BO2_lightbotmax_bdmax: *Light at bottom (maximum at max depth)*
* Geometry: the geometry in well-known text (wkt)
* Coastdist: *distance to the coast in meters*

	

***

